# Data

## Description
The dataset used in this project is the Startup Investments dataset documented by Crunchbase, publicly available on Github (https://raw.githubusercontent.com/notpeter/crunchbase-data/master/investments.csv), containing venture-backed startup information originally compiled from the Crunchbase platform and redistributed through an open data repository. The dataset represents a snapshot rather than a continuously updated feed and includes firm-level records collected by Crunchbase through self-reporting by companies, investor disclosures, and public announcements. The CSV file contains approximately 50,000 startups and 40 variables, including founding year, market category, geographic location, funding totals, number of funding rounds, and final company status (e.g., operating, acquired, closed, IPO). Each row corresponds to a single startup, with funding information aggregated across its lifetime rather than reported at a transactional level. Because the data are a static snapshot, they do not reflect real-time updates, and younger companies are more likely to appear as “operating,” introducing survivorship and censoring effects. Additional data quality issues include missing values in founding dates and funding fields, inconsistent market labels, and the absence of small or unreported startups that never appear in venture databases.

## Missing value analysis

```{r}
library(tidyverse)
library(janitor)
library(naniar)  # for missingness plots

startup_raw <- read_csv(
  "investments_VC.csv",
  locale = locale(encoding = "ISO-8859-1")   # matches Crunchbase examples
)

startup <- startup_raw |>
  clean_names() |>                        # makes names snake_case, trims spaces
  select(-starts_with("unnamed_")) |>     # drop completely empty columns
  distinct()                               # remove exact duplicates if any

glimpse(startup)
```

```{r}
status_palette <- c(
  operating = "#4C72B0",
  acquired  = "#55A868",
  closed    = "#C44E52",
  ipo       = "#8172B2"
)
```

```{r}
gg_miss_fct(startup, fct = status) +
  scale_fill_gradient(
    low  = "#DCE6F2",
    high = "#4C72B0"
  ) +
  labs(
    title    = "Missing Values by Startup Status",
    subtitle = "Count of missing entries across variables, grouped by status",
    x = "Startup status",
    y = "Number of missing values"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(size = 12),
    axis.text.x     = element_text(size = 9),
    axis.text.y     = element_text(size = 9),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )
```

The figure above summarizes the total number of missing entries across all variables by startup status, showing that operating startups account for the largest volume of missing data. 

Two columns (x41 and x40) in the raw file are entirely missing and are dropped. Location and founding information have substantial missingness (e.g., state_code, city, founded_year, founded_month, founded_quarter). Core variables for our analysis, including status, funding_total_usd, and name, are mostly observed, with only a small fraction of missing status (~2–3% of rows).

```{r}
founded_missing_by_status <- startup |>
  group_by(status) |>
  summarise(
    prop_missing_founded_year = mean(is.na(founded_year)),
    n = n(),
    .groups = "drop"
  )

gg_miss_status <- ggplot(
  founded_missing_by_status,
  aes(x = status, y = prop_missing_founded_year, fill = status)
) +
  geom_col(
    width = 0.7,
    alpha = 0.9
  ) +
  scale_fill_manual(values = status_palette) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1)
  ) +
  labs(
    title    = "Missing Founded Year Is Similar Across Startup Outcomes",
    subtitle = "Share of startups with missing founded_year within each status category",
    x = NULL,
    y = "Missing founded_year (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 12),
    axis.text.x   = element_text(size = 9),
    axis.text.y   = element_text(size = 9),
    panel.grid.minor = element_blank(),
    legend.position = "none"  
  )

gg_miss_status

```
This figure focuses specifically on the founding year variable and shows that the proportion of startups with missing founding dates is relatively similar across outcome categories, ranging from roughly 20% to 25% for acquired, closed, and operating firms. Although startups with missing or unknown status show slightly higher missingness about 30%, the overall similarity across outcomes suggests that missing founding-year information is unlikely to bias comparisons of outcomes over time. 

Together, these results indicate that while missing data are common—particularly in funding-related fields—their distribution is largely consistent across status categories, allowing downstream exploratory analyses to remain interpretable with appropriate filtering and caution.
