# Data

## Description

## Missing value analysis

```{r}
# install.packages("naniar")
```


```{r}
library(tidyverse)
library(janitor)
library(naniar)  # for missingness plots (optional but very handy)

startup_raw <- read_csv(
  "investments_VC.csv",
  locale = locale(encoding = "ISO-8859-1")   # matches Crunchbase examples
)

startup <- startup_raw |>
  clean_names() |>                        # makes names snake_case, trims spaces
  select(-starts_with("unnamed_")) |>     # drop completely empty columns
  distinct()                               # remove exact duplicates if any

glimpse(startup)
```

```{r}
# Long-format missingness summary
miss_summary <- startup %>%
  summarise(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "prop_missing") |>
  arrange(desc(prop_missing))

ggplot(miss_summary,
       aes(x = reorder(variable, prop_missing),
           y = prop_missing)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Proportion of Missing Values by Variable",
    x = NULL,
    y = "Missing values (%)"
  ) +
  theme_minimal()

```
Commentary you can write:

Two columns in the raw file are entirely missing and are dropped.

Location and founding information have substantial missingness (e.g., state_code, city, founded_year, founded_month, founded_quarter).

Core variables for our analysis – status, funding_total_usd, and name – are mostly observed, with only a small fraction of missing status (~2–3% of rows).

```{r}
gg_miss_fct(startup, fct = status) +
  labs(
    title = "Missingness by Startup Status",
    x = "Status",
    y = "Number of missing cells"
  )
```

```{r}
founded_missing_by_status <- startup |>
  filter(!is.na(status)) |>
  group_by(status) |>
  summarise(
    prop_missing_founded_year = mean(is.na(founded_year))
  )

ggplot(founded_missing_by_status,
       aes(x = status,
           y = prop_missing_founded_year)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Missing Founded Year by Startup Status",
    x = "Status",
    y = "% missing founded_year"
  ) +
  theme_minimal()

```

Commentary you can write:

Missingness in founding dates is fairly similar across operating, acquired, and closed startups, suggesting we do not have an obvious pattern where, for example, failed startups systematically omit founding information.

The status variable itself has a small proportion of missing values; these “unknown” cases could either be dropped in sensitivity analyses or treated as a separate category depending on the question.
