# Data

## Description
The dataset used in this project is the Startup Investments dataset documented by Crunchbase, publicly available on Github (https://raw.githubusercontent.com/notpeter/crunchbase-data/master/investments.csv), containing venture-backed startup information originally compiled from the Crunchbase platform and redistributed through an open data repository. The dataset represents a snapshot rather than a continuously updated feed and includes firm-level records collected by Crunchbase through self-reporting by companies, investor disclosures, and public announcements. The CSV file contains approximately 49,000 startups and 40 variables, including founding year, market category, geographic location, funding totals, number of funding rounds, and final company status (e.g., operating, acquired, closed, IPO). Each row corresponds to a single startup, with funding information aggregated across its lifetime rather than reported at a transactional level. Because the data are a static snapshot, they do not reflect real-time updates, and younger companies are more likely to appear as “operating,” introducing survivorship and censoring effects. Additional data quality issues include missing values in founding dates and funding fields, inconsistent market labels, and the absence of small or unreported startups that never appear in venture databases.

## Missing value analysis

```{r}
# install.packages("naniar")
```


```{r}
library(tidyverse)
library(janitor)
library(naniar)  # for missingness plots (optional but very handy)

startup_raw <- read_csv(
  "investments_VC.csv",
  locale = locale(encoding = "ISO-8859-1")   # matches Crunchbase examples
)

startup <- startup_raw |>
  clean_names() |>                        # makes names snake_case, trims spaces
  select(-starts_with("unnamed_")) |>     # drop completely empty columns
  distinct()                               # remove exact duplicates if any

glimpse(startup)
```

```{r}
miss_summary <- startup |>
  summarise(across(everything(), ~ mean(is.na(.)))) |>
  pivot_longer(everything(),
               names_to = "variable",
               values_to = "prop_missing") |>
  filter(prop_missing > 0) |>             # drop variables with no missing (optional)
  arrange(desc(prop_missing))

gg_miss_vars <- ggplot(miss_summary,
       aes(x = reorder(variable, prop_missing),
           y = prop_missing)) +
  geom_col(fill = "#4C72B0") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title    = "Some Variables Have Substantial Missing Data",
    subtitle = "Share of observations with missing values by variable",
    x = NULL,
    y = "Missing values (%)"
  ) +
  theme(
    plot.title = element_text(face = "bold")
  )

gg_miss_vars
```
Commentary you can write:

Two columns in the raw file are entirely missing and are dropped.

Location and founding information have substantial missingness (e.g., state_code, city, founded_year, founded_month, founded_quarter).

Core variables for our analysis – status, funding_total_usd, and name – are mostly observed, with only a small fraction of missing status (~2–3% of rows).

```{r}
gg_miss_fct(startup, fct = status) +
  labs(
    title = "Missingness by Startup Status",
    x = "Status",
    y = "Number of missing cells"
  )
```

```{r}
founded_missing_by_status <- startup |>
  group_by(status) |>
  summarise(
    prop_missing_founded_year = mean(is.na(founded_year)),
    n = n(),
    .groups = "drop"
  )

gg_miss_status <- ggplot(founded_missing_by_status,
       aes(x = status,
           y = prop_missing_founded_year)) +
  geom_col(fill = "#4C72B0") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title    = "Missing Founded Year Is Similar Across Startup Outcomes",
    subtitle = "Share of startups with missing founded_year within each status category",
    x = NULL,
    y = "Missing founded_year (%)"
  ) +
  theme(
    plot.title = element_text(face = "bold")
  )

gg_miss_status
```

Commentary you can write:

Missingness in founding dates is fairly similar across operating, acquired, and closed startups, suggesting we do not have an obvious pattern where, for example, failed startups systematically omit founding information.

The status variable itself has a small proportion of missing values; these “unknown” cases could either be dropped in sensitivity analyses or treated as a separate category depending on the question.
